---
title: "Assignment 3: Data Exploration"
author: "Shubhangi Gupta"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
getwd()
library(tidyverse)
library(lubridate)
Neonics <- read.csv("Data/Raw/Neonics.csv", stringsAsFactors = TRUE)
Litter <- read.csv("Data/Raw/Litter.csv", stringsAsFactors = TRUE)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Neonicotinoids are are a class of neuro-active insecticides chemically similar to nicotine developed by scientists at Shell and Bayer in the 1980s. Because they affect the central nervous system of insects, neonicotinoids kill or deleteriously affect a wide variety of both target and non-target insects. Their use has been linked to adverse ecological effects, including risks to many non-target organisms, and specifically on bees and pollinators. In 2022 the EPA concluded that neonicotinoids are likely to adversely affect the majority of federally listed endangered or threatened species and of critical habitats.They also widely contaminate wetlands, streams, and rivers, and due to their widespread use, pollinating insects are chronically exposed to them. Thus it is important to understand their ecotoxicology in order to understand if and how they adversely impact insects that could have cascading effects on our ecosystems and food systems. 
Source: Wikipedia (https://en.wikipedia.org/wiki/Neonicotinoid)

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter is material that dropped from the forest canopy and has a butt end diameter < 2cm and a length < 50 cm, while fine woody debris is dead, woody, plant material, including logs, branches, standing dead trees, and root wads that dropped from the forest canopy and have a butt end diameter < 2cm and a length >50. We are interested in studying them because woody debris is an important part of forest and stream ecosystems as it has a role in carbon budgets and nutrient cycling, is a source of energy for aquatic ecosystems, provides habitat for terrestrial and aquatic organisms, and contributes to structure and roughness, thereby influencing water flows and sediment transport (Harmon and others 1986). 
Source: (https://data.neonscience.org/data-products/DP1.10033.001
https://data.neonscience.org/data-products/DP1.10033.001)

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and fine woody debris are collected from elevated and ground traps, respectively.
 2.All masses reported following processing are reported at the spatial reso‐
lution of a single trap and the temporal resolution of a single collection event.
 3. Litter and fine woody debris sampling is executed at terrestrial NEON sites that contain woody vegetation > 2m tall. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics$Effect)
```

> Answer: The effects column gives information on how the Neonics chemicals impact the samples (mortality/ growth/ behavior/ reproduction/ etc). This is important to study to assess if the chemicals have a detrimental effect on the samples and of what kind. The most important effects are mortality and population in terms of data results as well as negative impact of the chemicals on insects.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}
head(sort(summary(Neonics$Species.Common.Name), decreasing=TRUE), 6)
```

> Answer:The most common species studied with their frequency in this dataset are: Other (670), (1)Honey Bee (667), Parasitic Wasp(285), Buff Tailed Bumblebee (183), Carniolan Honey Bee (152), Bumble Bee (140). What they have in common is that they are all different species of bees. They are important because they are the most important pollinating insects.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
```

> Answer: The class of this column is factor because data imported from CSV files is not always recognized as numeric data, especially if the data is not cleaned. In this case, the numeric data is accompanied with characters like /, ~ and >, rendering it a character vector instead of numeric.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
library(ggplot2)

ggplot(Neonics)+
   geom_freqpoly(aes(x=Publication.Year), col = "red", bins = 40)+
  ggtitle("Number of studies published each year")+
  labs (x = "Publication Year", y = "Number of Publications")
 
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics)+
  geom_freqpoly(aes(x=Publication.Year, color = Test.Location))+
  ggtitle("Number of studies published each year by Location")+
  labs (x = "Publication Year", y = "Number of Publications")
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test location between 1990-2000 and 2005-2010 was "field natural" while between 2000-2005 and 2010-2015 was the lab.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics)+
  geom_bar(aes(x=Endpoint), fill = "blue")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ggtitle("Endpoint Count")+
  labs (x = "Endpoint Category", y = "Number of Endpoints")

#Using code to double check the two most common endpoints
sort(table(Neonics$Endpoint), decreasing = TRUE)

```

> Answer: "NOEL" and "LOEL" are the two most common endpoints. The “highest tested concentration having no statistically significant adverse effect” is coded as NOEL and the “lowest tested concentration having a statistically significant adverse effect” is coded as a LOEL. Source: https://cfpub.epa.gov/ecotox/pdf/ecotoxsop.pdf

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
head(Litter$collectDate)
class(Litter$collectDate)
#The column collectDate is a factor.
Litter$collectDate <- ymd(Litter$collectDate)
class(Litter$collectDate)
unique(Litter$collectDate)
#The litter was collected on 2nd August 2018 and 30th August 2018.
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$uid)
summary(Litter)
```

> Answer:188 plots were sampled at Niwot Ridge
'Unique' gives the list of unique data in the column, whereas 'Summary' gives the number of datapoints in the column along with their class and mode. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter)+
  geom_bar(aes(x=functionalGroup), fill = "green") +
  ggtitle("Quantity of litter at the Niwot Ridge Sites")+
  labs (x = "Litter Categories", y = "Quantity of Litter")
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter)+
  geom_boxplot(aes(x=functionalGroup, y=dryMass), col = "purple")+
  ggtitle("Distribution of quantity of dry mass by Litter type")+
  labs (x = "Litter Type", y = "Quantile")

ggplot(Litter)+
  geom_violin(aes(x=functionalGroup, y=dryMass),
  draw_quantiles = c(0.25, 0.5, 0.75 ), col = "orange")+
  ggtitle("Distribution of quantity of dry mass by Litter type")+
  labs (x = "Litter Type", y = "Quantile")
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot shows the distribution of the data a lot more clearly along with the outliers whereas the violin plot shows the probability distribution of the data. In this case, given the outliers and low frquency of the data, the violin plot does not show much making it a less effective visualisation.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles tend to have the highest biomass at these sites.
