---
title: "Assignment 10: Data Scraping"
author: "Shubhangi Gupta"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
#install.packages("rvest")
library(rvest)
library(lubridate)
#install.packages("purrr")
library(purrr)

getwd()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
website = read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023")
website
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
WaterSystemName <- website%>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)")%>%
  html_text()
WaterSystemName

PWSID <- website %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)")%>%
  html_text()
PWSID

Ownership <- website %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)")%>%
  html_text()
Ownership

MGD <- website %>%
  html_nodes("th~ td+ td")%>%
  html_text()
MGD
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2023

```{r create.a.dataframe.from.scraped.data}
#4 
Month <- c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar","Jul", "Nov", "Apr", "Aug", "Dec")
Year <- rep(2023,12)
Date <- paste(Month, Year)

DurhamWaterSupply <- data.frame("Month" = Month, "City" = rep(WaterSystemName, 12), 
                                "PWSID" = rep(PWSID, 12), "Ownership" = rep(Ownership, 12), 
                                "MaxDayUse" = as.numeric(MGD), "Date" = my(Date))
glimpse(DurhamWaterSupply)
DurhamWaterSupply

#5 
ggplot(DurhamWaterSupply, aes(x=Date, y=MaxDayUse))+
  geom_line(col = "blue")+
  labs(title = paste("2023 Max Daily Withdrawals"),
       subtitle = WaterSystemName,
       y="Withdrawal (mgd)",
       x="Date")
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape.it <- function(year, the_PWSID){
  webpageURL <- read_html(paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?", 
                                 "pwsid=", the_PWSID,"&year=", year)) 
  
WaterSystemName <- webpageURL%>%html_nodes("div+ table tr:nth-child(1) td:nth-child(2)")%>% html_text()
PWSID <- webpageURL%>% html_nodes("td tr:nth-child(1) td:nth-child(5)")%>% html_text()
Ownership <- webpageURL%>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)")%>% html_text()
MGD <- webpageURL%>% html_nodes("th~ td+ td")%>% html_text()
Month <- c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar","Jul", "Nov", "Apr", "Aug", "Dec")
Year <- rep(year,12)
Date <- paste(Month, Year)

WaterSupplyFunction <- data.frame("Month" = Month, "City" = rep(WaterSystemName, 12), "PWSID" = rep(PWSID, 12), "Ownership" = rep(Ownership, 12), "MaxDayUse" = as.numeric(MGD), "Date" = my(Date))

return(WaterSupplyFunction)
}
```



7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
Durham2015 <- scrape.it(2015,'03-32-010')
Durham2015

ggplot(Durham2015, aes(x=Date, y=MaxDayUse))+
  geom_line(col = "blue")+
  labs(title = paste("2015 Max Daily Withdrawals"),
       subtitle = WaterSystemName,
       y="Withdrawal (mgd)",
       x="2015")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
Asheville2015 <- scrape.it(2015, "01-11-010")
Asheville2015

ggplot()+
  geom_line(data=Durham2015, aes(x=Date, y=MaxDayUse), col="blue")+
  geom_line(data=Asheville2015, aes(x=Date, y=MaxDayUse), col="green")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
AshevilleYears <- rep(2010:2021)
AshevilleThroughTheYears <- map2(AshevilleYears, "01-11-010", scrape.it )
AshevilleThroughTheYears
AshevilleThroughTheYears_All <- bind_rows(AshevilleThroughTheYears)

#Plot
ggplot(AshevilleThroughTheYears_All,aes(x=Date,y=MaxDayUse)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("Max Daily Withdrawals from 2010-21"),
       subtitle = WaterSystemName,
       y="Withdrawal (mgd)",
       x="2015")

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: Kind of. There is a lot of fluctuation in the data with a possible seasonal component. Until 2015 there was no significant trend in water usage, but since then it seems to have started to rise, albeit with a lot of fluctuation.
